\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-30 tex/paper.tex
save

Complex computer simulators are used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
continuous or discrete observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
Preliminary results with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
validate the proposed method.

@@

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which may result
in learning a smeared variation of the original simulator. On the other hand,
for large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. A too large penalty may however
eventually make the optimization unstable, as the variance of $\nabla_\psi \log q(\theta_m|\psi)$
typically increases as the entropy of the proposal decreases.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Likelihood $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again
slightly better results for the entropy penalized model. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

Closest to our work are procedures that rely on a classifier to estimate the
discrepancy between the true and the model distributions. For example,
\citep{gutmann2012noise} uses non linear logistic regression for fitting
unnormalized differentiable statistical models, while
\citep{goodfellow2014generative} exploits an adversarial neural network for
learning a differentiable implicit generative model. In the likelihood-free
setup, \citep{cranmer2015approximating,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
parameter inference in combination with a gradient-free optimization algorithm.
Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.

In this context, the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative} to
non-differentiable simulators. It also constitutes an approximate gradient
descent alternative to bayesian optimization based on density
ratios~\citep{cranmer2015approximating} or to classifier
ABC~\citep{gutmann2017likelihood}.

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
continuous or discrete observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
Preliminary results with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
validate the proposed method.

@@

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which may result
in learning a smeared variation of the original simulator. On the other hand,
for large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. A too large penalty may however
eventually make the optimization unstable, as the variance of $\nabla_\psi \log q(\theta_m|\psi)$
typically increases as the entropy of the proposal decreases.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
adversarial variational optimization. We see that the distributions both arrive
in the neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\theta^*$ or that the observed data can in this case also be
reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of the simulator.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution. In summary, despite the intricate
relation betwen the observations $\mathbf{x}$, the latent variable $\mathbf{z}$
and the parameters $\theta$, this example demonstrates the efficacy
of adversarial variational optimization for parameter inference on a black box
simulator.

@@

Closest to our work are procedures that rely on a classifier to estimate the
discrepancy between the true and the model distributions. For example,
\citep{gutmann2012noise} uses non linear logistic regression for fitting
unnormalized differentiable statistical models, while
\citep{goodfellow2014generative} exploits an adversarial neural network for
learning a differentiable implicit generative model. In the likelihood-free
setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
parameter inference in combination with a gradient-free optimization algorithm.
Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.

In this context, the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative} to
non-differentiable simulators. It also constitutes an approximate gradient
descent alternative to bayesian optimization based on density
ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier
ABC~\citep{gutmann2017likelihood}.
